---
title: "DEGAIN TRAINING 2026 - Ancestral State Reconstruction"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    theme: cosmo
    highlight: textmate
    number_sections: true
editor_options: 
  chunk_output_type: inline
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, eval=FALSE, fig.width=8, fig.height=8)
```


## Introduction & Prerequisites

Welcome to the **ASR Training Workbook**. This guide contains code and instructions to perform ancestral state reconstruction for discrete bacterial traits of interest using dated phylogenies.

We assume you have already run **BactDating** and have your time-calibrated tree ready.

### Learning Objectives

By the end of this workbook, you will be able to:

1. Load and visualise dated phylogenetic trees and metadata in R.
2. Perform ancestral state reconstruction of discrete traits.


### Directory Setup

Ensure your workspace is organized as follows:
```text
project_root/
├── data/data_kpn/
│   ├── dated_tree.tree
│   └── metadata.tsv        # metadata with traits info
└── scripts/
└── asr.Rmd          # This workbook
```

---
    
    
```{r}
library(tidyverse)
library(ggtree)
library(treeio)
library(phytools)

source("scripts/asr_fnxs.R")
```

```{r}
seed <- 98437621
```


```{r load_data}
# Set data directories
project_dir <- here::here()
my_data_dir <- file.path(project_dir, "data/data_kpn/") 

# Load metadata
metadata <- read_tsv(file.path(my_data_dir, "metadata.tsv"), show_col_types=F) 

# Load dated tree
tree_data <- treeio::read.beast(file.path(my_data_dir, "dated_tree.tree"))
tree <- tree_data@phylo

# Set 0 branch lengths to very small num (ace doesnt like these)
tree$edge.length[tree$edge.length==0]<-max(nodeHeights(tree))*1e-6

# Subset metadata based on tree tips and get columns for asr
tree_meta <- metadata %>% slice(match(tree$tip.label, sequence_id)) %>% 
    select(sequence_id, region, country, K_locus, O_type, year) 

# join metadata to tree
tree_and_clone_data <- full_join(tree_data, tree_meta %>% rename("label"="sequence_id"), 
                                 by = "label")

# mrsd
data_mrsd_year <- max(tree_meta$year, na.rm=T)
data_mrsd <- paste0(data_mrsd_year, "-01-01")
data_mrsd_rtt <- max(nodeHeights(tree)) # longest root to tip distance (years)

# make k locus vec for asr 
k_loci_data <- tree_meta %>% select(sequence_id, K_locus) %>% deframe() %>% as.factor()

# make country data for asr
country_data <- tree_meta %>% select(sequence_id, country) %>% deframe() %>% as.factor()

# get k locus colours for subsequent plot(s)
k_loci_cols <- get_colours(k_loci_data)

# get country colours for subsequent plot(s)
country_cols <- get_colours(country_data)
```


```{r k_asr}
# Use the equal rates model and stochastic character mapping method
n_simulations <- 100
set.seed(seed)
ktrees <- phytools::make.simmap(tree, k_loci_data, model="ER", nsim = n_simulations)

# pick a representative tree
k_rep_tree <- ktrees[[1]]

# summarise stochastic character maps
k_simmap_summary <- summary(ktrees, ref.tree = tree)
print(k_simmap_summary)

# plot summary on tree
plot(k_simmap_summary, colors=k_loci_cols, fsize=0.000000001)

# more detailed summary
k_described_simmap <- phytools::describe.simmap(ktrees)

# calculate mean time spent in each state
k_mean_times <- apply(k_described_simmap$times, 2, mean) %>%
    as.list() %>% 
    data.frame(check.names=FALSE) %>% 
    as_tibble() %>% 
    pivot_longer(cols = c(-total), names_to = "locus", values_to = "years") %>% 
    mutate(prop_of_time_in_state = years/total) %>% 
    select(locus, total_tree_time = total, years_in_state = years,
           prop_of_time_in_state)

# calculate mean transitions
k_mean_counts <- apply(k_described_simmap$count, 2, mean) %>%
    as.list() %>% 
    data.frame(check.names=FALSE) %>% 
    as_tibble() %>%
    rename("mean_transitions" = "N")

# plot mean transitions
k_mean_counts_plot <- k_mean_counts %>% 
    pivot_longer(-c(mean_transitions), values_to = "transitions") %>% 
    mutate(transitions = round(transitions, 2)) %>% 
    separate_wider_delim(name, delim=",", names = c("loc1", "loc2")) %>% 
    ggplot(aes(x=loc1, y=loc2, fill=transitions)) +
    geom_tile() + 
    geom_text(aes(label = transitions), color = "black", size = 3) +
    theme_classic() + 
    scale_fill_distiller(palette = "YlOrRd", name = "Transitions") + 
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
          axis.text = element_text(size = 12),
          legend.title = element_text(size = 12)) + 
    labs(title = "K transition matrix", x=NULL, y=NULL)
k_mean_counts_plot
ggsave(file.path(my_data_dir, "k_transitions.png"),
       plot=k_mean_counts_plot, height=7, width=8, bg="white")

# plot reconstructed tree
png(file.path(my_data_dir, "k_asr_tree.png"),
    width = 280, height = 325, units='mm', res = 300)
plot_ancestral_state(k_rep_tree, k_loci_data, k_simmap_summary,
                     colours=get_colours(k_loci_data), tree_type='phylogram',
                     add_node_pies=T, legend_size=0.9)
dev.off()
```

### Discussion

1. In the code, we run `n_simulations <- 100 models`. Why do we simulate 100 character maps instead of just running a single reconstruction? What statistical advantage does this pool of simulations provide when we call `summary(ktrees)`?

2. When rendering the final plot, the function uses `k_rep_tree <- ktrees[1]`. What is this tree? What is the difference between what `ktrees[1]` represents versus what `ksimmap_summary$ace` represents?

3. We specified `model="ER"` in `make.simmap()`. What does this mean? In what biological scenarios would an 'ER' model likely be flawed?

4. Look closely at how `k_mean_times` and `k_mean_counts` are calculated. If a tree shows that a lineage spent 80% of its total evolutionary time in a specific state, and the transition matrix shows zero transitions originating from that state, how would you interpret the evolutionary stability of that trait (not just the specific state in question)? In our example, what can you say about the stability of the K loci in this lineage?

5. Look at the pie charts plotted on the nodes of your reconstructed ancestral state tree. If a node is split 50/50 between two colors, what does this tell you about the ancestral state at that specific point in evolutionary time? How should we talk about this node in our results compared to a node that is 100% one color?

6. Examine the k transition matrix heatmap. Which transitions appear to be the most frequent?


---

### Further exercise

Using the code above, reconstruct ancestral states of the country of isolation 
throughout the tree and prepare a 2-slide presentation to discuss any interesting findings with the class. 

```{r country_asr}
# Use the equal rates model and stochastic character mapping method
n_simulations <- 100
set.seed(seed)
ctrytrees <- phytools::make.simmap(tree, country_data, model="ER", nsim = n_simulations)

# pick a representative tree
ctry_rep_tree <- ctrytrees[[1]]

# summarise stochastic character maps
ctry_simmap_summary <- summary(ctrytrees, ref.tree = tree)
print(ctry_simmap_summary)

# plot summary on tree
plot(ctry_simmap_summary, colors=country_cols, fsize=0.000000001)

# more detailed summary
ctry_described_simmap <- phytools::describe.simmap(ctrytrees)

# calculate mean time spent in each state
ctry_mean_times <- apply(ctry_described_simmap$times, 2, mean) %>%
    as.list() %>% 
    data.frame(check.names=FALSE) %>% 
    as_tibble() %>% 
    pivot_longer(cols = c(-total), names_to = "locus", values_to = "years") %>% 
    mutate(prop_of_time_in_state = years/total) %>% 
    select(locus, total_tree_time = total, years_in_state = years,
           prop_of_time_in_state)

# calculate mean transitions
ctry_mean_counts <- apply(ctry_described_simmap$count, 2, mean) %>%
    as.list() %>% 
    data.frame(check.names=FALSE) %>% 
    as_tibble() %>%
    rename("mean_transitions" = "N")

# plot mean transitions
ctry_mean_counts_plot <- ctry_mean_counts %>% 
    pivot_longer(-c(mean_transitions), values_to = "transitions") %>% 
    mutate(transitions = round(transitions, 2)) %>% 
    separate_wider_delim(name, delim=",", names = c("loc1", "loc2")) %>% 
    ggplot(aes(x=loc1, y=loc2, fill=transitions)) +
    geom_tile() + 
    geom_text(aes(label = transitions), color = "black", size = 3) +
    theme_classic() + 
    scale_fill_distiller(palette = "YlOrRd", name = "Transitions") + 
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
          axis.text = element_text(size = 12),
          legend.title = element_text(size = 12)) + 
    labs(title = "Country transition matrix", x=NULL, y=NULL)
ctry_mean_counts_plot
ggsave(file.path(my_data_dir, "ctry_transitions.png"),
       plot=ctry_mean_counts_plot, height=7, width=8, bg="white")

# plot reconstructed tree
png(file.path(my_data_dir, "ctry_asr_tree.png"),
    width = 280, height = 325, units='mm', res = 300)
plot_ancestral_state(ctry_rep_tree, country_data, ctry_simmap_summary,
                     colours=get_colours(country_data), tree_type='phylogram',
                     add_node_pies=T, legend_size=0.9)
dev.off()

```


